\normalsize{A general approach and tool to evaluate the effectiveness of no-take marine reserves}

### 2017-08-21 Manuscript.Rmd
Updated manuscript

Marine reserves are often implemented to preserve habitat, recover overfished stocks, and secure livelihoods of coastal communities. As with any other management interventions, we need to better understand their effectiveness and impacts on coastal communities and the environment. To date, the evaluation of these reserves largely relies on analyzing ecological data, often ignoring socioeconomic and governance dimensions. Existing data are analyzed in different ways, hindering the ability to compare results across case studies. Moreover, analysis and evaluation of reserves is generally conducted by academic scientists, not the reserves managers and users, thereby hindering effective local management and rapid response to change. We present a framework and tool to evaluate the effectiveness of marine reserves by matching seven commonly stated management objectives to nine biological, five socioeconomic, and 14 governance indicators. We provide guidelines on how to properly collect data that can then be analyzed with standardized method. Biological indicators are evaluated with causal inference techniques, using a counterfactual approach, to assess the net effect of the reserve on each indicator. Linear regression models are fitted to socioeconomic indicators through time to test for differences before and after reserve implementation. Governance indicators are qualitatively analyzed using a framework developed through a literature review which identifies common governance structures and their associated effectiveness. To make the framework accessible to fishers and decision makers, and allow replication of results, we developed the open source, web-based Marine Reserve Evaluation App (MAREA). Together, this new framework and MAREA can further our understanding and support management of marine reserves.

@@

Overfishing and unsustainable fishing practices are two of the largest threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK]. Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish and invertebrate stocks recover [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limiting or restricting fishing effort and gears. No-take marine reserves (marine reserves hereinafter) are a particular type of MPA, where all fishing effort and extractive activities are off-limits [-@iucn_website_2017-Bg; @hortaecosta_2016-_3].

Empirical studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg]. Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO]. These effects are often measured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq]. This before-after comparison cannot account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves. While some studies have used control sites, these analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD]. A smaller fraction of studies have used a before-after-control-impact design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.

As with any other policy intervention, it is important that we can measure its effect in order to adapt and learn [@ferraro_2006-oW; @burgess_2016-HN]. The diversity of approaches currently used to evaluate the effect of marine reserves often does not answer the simple question: *What is the effect of a marine reserve on a given attribute?* This gap highlights the need to develop standardized approaches that enable us to evaluate the net effect of the intervention (*i.e.* causes of conservation outcomes; @ferraro_2006-oW). Furthermore, while biological aspects are important to reserve success, effectiveness also depends on the socioeconomic status and governance system of the local fishing communities [@basurto_2013-oq; @basurto_2012-Mz], which are often ignored. By excluding these important dimensions, the evaluation provides only a partial picture of the impacts of the reserve. Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these indicators may be measured, but does not provide guidelines on how to analyze them. Recent work by @mascia_2017-m_ integrates these three dimensions and suggests the use of causal inference techniques to provide a measure of the effect of implementing an MPA. However, these two novel approaches do not provide a user-friendly tool that enables replicability and scalability of the analysis, particularly when used by the fishers and decision makers themselves.

An increasingly popular way to make science reproducible, scalable and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh]. The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has successfully standardized a way to measure the health and benefits of the oceans. This approach has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC]. Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydropower industry [@vilela_2017-Zo], just to list a few.

The lack of a comprehensive framework and user-friendly tools to evaluate the effectiveness of marine reserves ---or the complexity of existing ones, which alienate non--experts--- calls for the development of a new framework and tool. The current work presents a framework to evaluate marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these areas. We first provide a list of commonly stated management objectives and match them to appropriate indicators that measure the effectiveness of the management intervention. We then include a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_]. We also introduce the Marine Reserve Evaluation App (MAREA), an open source web--based tool that automates the framework described in this paper and enables its broader use. Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.

@@

Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (Fig. 1). We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection. Alongside, methodologies to analyze these indicators are presented. Then, we describe the development of MAREA and explain how this user-friendly open access tool can be used by anyone. Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA.

![Workflow to evaluate the effectiveness of marine reserves.](./Figures/flowchart.gif){ width=50% }

Throughout this study, we will refer to the stated goals for which a marine reserve was designed as objectives. This work was motivated by the need to provide a framework to evaluate Mexican marine reserves. Thus, our focus was on identifying common objectives of marine reserves in Mexico. However, we group these objectives into seven major categories, which can be applied to marine reserves worldwide. The list of objectives was developed through a literature review, which compiled stated objectives in and legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, and decrees associated to these areas. Even though each reserve has its own goals, seven main categories of objectives were identified:

1.	Avoid overexploitation
2.	Conserve species under a special protection regime
3.	Maintain biological process (reproduction, recruitment, growth, feeding)
4.	Improve fishery production in nearby waters
5.	Preserve biological diversity and the ecosystem
6.	Recover overexploited species
7.	Recover species of economic interest

Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness. The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives. A first filter eliminated indicators for which baseline data do not typically exist in Mexico. The preliminary list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations. Later, these were also presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P. Ensenada*), in Baja California, who provided input. Our final list of indicators includes those identified in review works such as @lester_2009-Ks or @woodcock_2017-Wm.

Indicators are divided into three main categories: biological, socioeconomic, and governance (Table 1). Biological indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies). Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches, and availability of alternative livelihoods. Governance indicators (n = 15) describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve). Indicators may be numeric (e.g. Fish biomass) or descriptive (*e.g.* Reasoning for reserve location). Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance. In that sense, most biological and socioeconomic indicators are outcome variables. On the other hand, governance indicators are viewed as possible explanatory variables of reserve performance. Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass). Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with indicators.

@@

```{r, results = "asis"}
read.csv("./Tables/Table_indicators.csv") %>% 
  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves, grouped by type. Type of data and units are provided.", col.names = c("Code", "Indicator", "Data type", "Unit"), booktabs = T) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  group_rows("Biophysical", 1, 7) %>%
  group_rows("Socioeconomic", 8, 10) %>% 
  group_rows("Governance", 11, 25) %>% 
  add_footnote(c("The indicator is applied to objective species"), notation = "symbol")
```

```{r}
read.csv("./Tables/Objectives_Indicators.csv") %>% 
  knitr::kable(caption = "This table indicates which indicators must be used for each objective.",
               col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1**", "S2", "S2*", "S3", "G1", "G2", "G3", "G4", "G5", "G6", "G7", "G8", "G9", "G0", "G1", "G2", "G3", "G4", "G5"),
               booktabs = T) %>% 
  kable_styling(latex_options = c("scale_down"))  %>% 
  column_spec(1, width = "4cm") %>%
  landscape()
```

@@

In many coastal marine reserves of Mexico, biological data are collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance from Civil Society Organizations (CSOs). Scientific divers record fish and invertebrate richness and abundances, as well as fish total length along belt transects. Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact (*i.e.* BACI) sampling design. Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity. While transect dimensions (i.e. length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study-specific standardizes way. For this reason, MAREA does not assume specific transect dimensions, and pertinent indicators are calculated per transect (Table 1).

Having control sites for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the net effect of the reserve. The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models [@moland_2013-VP]. To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into a model with form:

@@

In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively. This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates treatment ($Z$; *i.e.* control or reserve), an interaction between a dummy variable that indicates pre-- or post--implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m). $\epsilon$ represents the error term associated to the equation. Here, years are modeled as factors, using the first year as the reference level. By modelling years as factors, we avoid imposing a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2006 and 2007 does not have to be the same as the change between 2015 and 2016). The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.

Socioeconomic data are often collected by fishers, fishery management agencies or OSCs by recording landings, income, and sometimes prices for each species. To control for inflation and changes in buying power, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:

@@

While this model does not allow establishing a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input. For both models (eq. 1 and eq. 3), coefficients are estimated via ordinary least squares, and heteroskedastic-robust standard errors are calculated.

Governance data are not readily available nor systematically collected by the community or other organizations. Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15). The survey is included as supplementary material (Appendix 1). To analyze governance information, we develop a framework based on a literature review of common governance structures and their relation to effectiveness in managing fisheries or marine reserves (Appendix 2). This approach has been proven to successfully evaluate governance structures [@espinosaromero_2014-PY]. Governance information is not quantitatively analyzed, but it is presented along with the biological and socioeconomic indicators to provide managers and users with a more complete description of the reserve.

@@

MAREA was developed in R Studio [@rcore_2017] using the Shiny package [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server. MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/). While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.

MAREA is designed as a 6-step process, divided in tabs. The first tab introduces the app and summarizes the evaluation process. Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2. Users can also manually select additional indicators or unselect the default ones, based on their interests and data availability. The user can then load the data, using standard \*.csv text files; sample datasets are provided within MAREA. Once data have been loaded, MAREA identifies all reserves in the data (the uploaded dataset can contain data for more than one reserve), and lets the user select the reserve to be evaluated. At this point, the user can also specify the year of implementation of the reserve, reserve dimensions, and indicate target species that are of particular management interest. Before presenting the results, MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct. Finally, the user is taken to the results tab where they can be viewed in a simple format. The user can also download a more comprehensive technical report produced in \*.pdf format.

The first output is a color-coded scorecard intended to provide a general overview of the effectiveness of the reserve. The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator. The global and category-level scores are determined by the percentage of positive indicators, overall and for each category, respectively. For numeric biological indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in eq. 1. For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$). Red, yellow and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively. The intensity of the color is defined by the significance of coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Students t-test. Cutoff values are $p < 0.05$ and $p < 0.1$. Thus, even in a case where $\beta_i > 0$, if the coefficient is not significant (*i.e. * $p>0.1$), the indicator will be assigned a yellow color. A legend (Fig. 2) is provided within the scorecard to aid in the interpretation of these results. Governance indicators are represented simply by red or green. The color is defined based on what literature shows to be a negative (red) or positive (green) factor for a reserve. For example, if the perceived degree of illegal fishing is high, this is indicator will be assigned a red color. However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the technical report).

![Legend used to interpret the scorecard produced by MAREA. Colors indicate direction of change, and color intensity is given by the statistical significance.](./Figures/legend2.gif)

The second output from MAREA is a technical report intended to communicate information and statistical results in a more comprehensive and technical way. This report also includes a scorecard as a summary of the results, but provides more information for each indicator. For all numeric indicators, the report includes a graph of the value of the indicator, for both the reserve and control sites, through time. It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors. The summary table also provides information on model fit ($R^2$) and significance of the regression.

The scorecard is produced with functions from the Shinydashboard package [@shinydashboard_2017]. The technical report is produced by a parameterized Rmarkdown document [@rmarkdown_2017] processed by the knitr package [@knitr_2017]. Another feature of MAREA is that the user can choose to share the data. Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data is saved into a central repository. These data can be accessed at any time by any person interested in acquiring them at [github.com/turfeffect/MAREAdata](github.com/turfeffect/MAREAdata).

@@

```{r map, fig.cap = "General location of Isla Natividad (left) and map of the island (right). The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce)."}
proj <- CRS("+proj=longlat +datum=NAD27")

@@

study_area <- tm_shape(baja2, bbox = bb(xlim = c(-115.22, -115.14), ylim = c(27.85, 27.90))) +
  tm_polygons() +
  tm_shape(reserves) +
  tm_polygons(border.col = "black", col = "red", border.lwd = 0) +
  tm_shape(controls) +
  tm_symbols(shape = 22, size = 4, border.col = "black", col = "blue", border.lwd = 1) +
  tm_text("Label") +
  tm_shape(labels2) +
  tm_text("Label") +
  tm_compass(position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom"))

tmap_arrange(pen_map, study_area, outer.margins = 0.005, ncol = 2)
```

@@

>>>>

Marine reserves are often implemented to preserve habitat, recover overfished stocks, and secure livelihoods of coastal communities. As with any other management interventions, we need to better understand their effectiveness and impacts on coastal communities and the environment. To date, their evaluation largely relies on analyzing ecological data, often ignoring socioeconomic and governance dimensions. Existing data are analyzed in different ways, hindering the ability to compare results across case studies. Moreover, analysis and evaluation of reserves is generally conducted by academic scientists, not the reserves managers and users, thereby hindering effective local management and rapid response to change. We present a framework and tool to evaluate the effectiveness of marine reserves by matching seven commonly stated management objectives to nine biological, five socioeconomic, and 15 governance indicators. We provide guidelines on how to properly collect data that can then be analyzed with standardized method. Biological indicators are evaluated with causal inference techniques, using a counterfactual approach, to assess the effect of the reserve on each indicator. Linear regression models are fitted to socioeconomic indicators through time to test for differences before and after reserve implementation. Governance indicators are qualitatively analyzed using a framework developed through a literature review, which identifies common governance structures and their associated effectiveness. To make the framework accessible to fishers and decision makers, and allow replication of results, we developed the open source, web-based Marine Reserve Evaluation App (MAREA). Together, this new framework and MAREA can further our understanding and support management of marine reserves.

@@

Overfishing and unsustainable fishing practices are two of the largest threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK]. Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish and invertebrate stocks recover [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV; @hastings_2017-sm] by limiting or restricting fishing effort and gears. No-take marine reserves (marine reserves hereinafter) are a particular type of MPA, where all fishing effort and extractive activities are off-limits [-@iucn_website_2017-Bg; @hortaecosta_2016-_3].

Empirical studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg]. Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO]. These effects are often measured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq]. This before--after comparison cannot account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves. While some studies have used control sites, these analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD]. A smaller fraction of studies have used a before-after-control-impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.

As with any other policy intervention, it is important that we can measure its effect in order to adapt and learn [@ferraro_2006-oW; @burgess_2016-HN]. The diversity of approaches currently used to evaluate the effect of marine reserves often does not answer the simple question: *What is the effect of a marine reserve on a given attribute?* This gap highlights the need to develop standardized approaches that enable us to evaluate the net effect of the intervention (*i.e.* causes of conservation outcomes; @ferraro_2006-oW). Furthermore, while biological aspects are important to reserve success, effectiveness also depends on the socioeconomic status and governance system of the local fishing communities [@basurto_2013-oq; @basurto_2012-Mz], which are often ignored. By excluding these important dimensions, the evaluation provides only a partial picture of the impacts of the reserve.

Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these indicators may be measured, but does not provide guidelines on how to analyze them. Recent work by @mascia_2017-m_ integrates these three dimensions and suggests the use of causal inference techniques to provide a measure of the effect of implementing an MPA. However, these two novel approaches do not provide a user-friendly tool that enables replicability and scalability of the analysis, particularly when used by the fishers and decision makers themselves.

An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh]. The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has successfully standardized a way to measure the health and benefits of the oceans. This approach has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC]. Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydro power industry [@vilela_2017-Zo], just to list a few.

The lack of a comprehensive framework and user-friendly tools to evaluate the effectiveness of marine reserves ---or the complexity of existing ones, which alienate non--experts--- calls for the development of a new framework and tool. The current work presents a framework to evaluate marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these areas. We first provide a list of commonly stated management objectives and match them to appropriate indicators. We then include a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_]. We also introduce the Marine Reserve Evaluation App (MAREA), an open source, web--based tool that automates the framework described in this paper and enables its broader use. Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.

Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (Fig. 1). We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection. Alongside, methodologies to analyze these indicators are presented. Then, we describe the development of MAREA and explain how this user-friendly open access tool can be used by anyone. Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.

**Fig. 1. Workflow to evaluate the effectiveness of marine reserves.**

```{r, fig.cap = "Workflow to evaluate the effectiveness of marine reserves."}
fig1 <- knitr::include_graphics("./Figures/flowchart.gif", dpi = 2380)
```

Throughout this study, we will refer to the stated goals for which a marine reserve was designed as objectives. This work was motivated by the need to provide a framework to evaluate Mexican marine reserves. Thus, our focus was on identifying common objectives of marine reserves in Mexico. However, we group these objectives into seven major categories, which can be applied to marine reserves worldwide. The list of objectives was developed through a literature review, which compiled stated objectives in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, and decrees associated to these areas. Even though each reserve has its own goals, seven main categories of objectives were identified:

1.	Avoid overexploitation
2.	Conserve species under a special protection regime
3.	Maintain biological processes (reproduction, recruitment, growth, feeding)
4.	Improve fishery production in nearby waters
5.	Preserve biological diversity and the ecosystem
6.	Recover overexploited species
7.	Recover species of economic interest

Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness. The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives. A first filter eliminated indicators for which baseline data do not typically exist in Mexico. The preliminary list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations. Later, these were presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P. Ensenada*), in Baja California, who provided input. Our final list of indicators includes those identified in review works such as @lester_2009-Ks or @woodcock_2017-Wm.

Indicators are divided into three main categories: biological, socioeconomic, and governance (Table 1). Biological indicators (n = 9) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies). Socioeconomic indicators (n = 5) reflect the performance of the fishery in terms of landings, income from landings, and availability of alternative livelihoods. Governance indicators (n = 15) describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve). Indicators may be numeric (e.g. Fish biomass) or descriptive (*e.g.* Reasoning for reserve location). Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance. In that sense, most biological and socioeconomic indicators are outcome variables. On the other hand, governance indicators are viewed as possible explanatory variables of reserve performance. Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass). Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators; all governance indicators should always be used.

@@

```{r, results = "asis"}
read.csv("./Tables/Table_indicators.csv") %>% 
  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves, grouped by type.",
               col.names = c("Code", "Indicator", "Data type", "Unit")) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  group_rows("Biological", 1, 7) %>%
  group_rows("Socioeconomic", 8, 10) %>% 
  group_rows("Governance", 11, 25) %>% 
  add_footnote(c("The indicator is applied to target species"), notation = "symbol")
```

@@

```{r}
read.csv("./Tables/Objectives_Indicators.csv") %>% 
  select(1:15) %>% 
  knitr::kable(caption = "Management objectives and respective performance indicators.",
               col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
  kable_styling(latex_options = c("scale_down"))  %>% 
  column_spec(1, width = "4cm") %>% 
  add_footnote(c("The indicator is applied to target species"), notation = "symbol")
```

@@

In many coastal marine reserves of Mexico, biological data are collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance from Civil Society Organizations (CSOs). Scientific divers record fish and invertebrate richness and abundances, as well as fish total length along belt transects. Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a BACI sampling design. Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity. While transect dimensions (*i.e.* length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study--specific standardized way. For this reason, MAREA does not assume specific transect dimensions, and pertinent indicators are calculated per transect (Table 1).

Having a BACI sampling design for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the effect of the reserve on biological indicators. The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models [@moland_2013-VP]. To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into difference-in-differences a model with form:

@@

In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively. This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates treatment ($Z$; *i.e.* control or reserve), an interaction between a dummy variable that indicates pre-- or post--implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m). $\epsilon$ represents the error term associated to the regression. Here, years are modeled as factors, using the first year as the reference level. This does not impose a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2006 and 2007 does not have to be the same as the change between 2015 and 2016). The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre--implementation = 0 and Post--implementation = 1, respectively.

Socioeconomic data are often collected by fishers, fishery management agencies or CSOs by recording landings, income, and sometimes prices for each species. To control for inflation, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:

@@

While this model does not allow establishing a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input. For both models (eq. 1 and eq. 3), coefficients are estimated via ordinary least squares, and heteroskedastic--robust standard errors are calculated.

Governance data are not readily available nor systematically collected by the community or other organizations. Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15). The survey is included as supplementary material in English (S1 Appendix) and Spanish (S2 Appendix). To analyze governance information, we developed a framework based on a literature review of common governance structures and their relation to effectiveness in managing fisheries or marine reserves (S3 Table). This approach has been proven to successfully evaluate governance structures [@espinosaromero_2014-PY]. Governance information is not quantitatively analyzed, but it is presented along with the biological and socioeconomic indicators to provide managers and users with a more complete description of the reserve.

@@

MAREA was developed in R version 3.4.1 and R Studio [@rcore_2017] using the Shiny package [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server. MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/). While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.

MAREA is designed as a 6-step process, divided in tabs. The first tab introduces the app and summarizes the evaluation process. Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2. Users can also manually modify selected indicators based on their interests and data availability. The user can then load the data, using standard \*.csv text files; sample datasets are provided within MAREA. Once data have been loaded, MAREA identifies all reserves in the data (the uploaded dataset can contain data for more than one reserve), and lets the user select the reserve to be evaluated. At this point, the user can also specify the year of implementation of the reserve, reserve dimensions, and indicate target species that are of particular management interest. Before presenting the results, MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct. Finally, the user is taken to the results tab where they can be viewed in a simple format. The user can also download a more comprehensive technical report produced in \*.pdf format.

The first output is a color--coded scorecard intended to provide a general overview of the effectiveness of the reserve. The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator. The global and category--level scores are determined by the percentage of positive indicators, overall and for each category, respectively. For numeric biological indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in eq. 1. For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$). Red, yellow and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively. The intensity of the color is defined by the significance of the coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Student's t-test. Cutoff values are $p < 0.05$ and $p < 0.1$. Thus, even in a case where $\beta_i > 0$, if the coefficient is not significant (*i.e. * $p>0.1$), the indicator will be assigned a yellow color. A legend (Fig. 2) is provided within the scorecard to aid in the interpretation of these results. Governance indicators are represented simply by red or green. The color is defined based on what literature shows to be a negative (red) or positive (green) factor for a reserve (S3 Table). For example, if the perceived degree of illegal fishing is high, this is indicator will be assigned a red color. However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the technical report).

**Fig. 2. Legend used to interpret the scorecard produced by MAREA.** Colors indicate direction of change (red = negative; green = positive), and color intensity is given by the statistical significance.

```{r}
fig2 <- knitr::include_graphics("./Figures/legend2.gif")
```

The second output from MAREA is a technical report intended to communicate information and statistical results in a more comprehensive and technical way. This report also includes a scorecard as a summary of the results, but provides more information for each indicator. For all numeric indicators, the report includes a graph of the value of the indicatorin the reserve and control sites, through time. It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors. The summary table also provides information on model fit ($R^2$) and significance of the regression.

The scorecard is produced with functions from the Shinydashboard package [@shinydashboard_2017]. The technical report is produced by a parameterized Rmarkdown document [@rmarkdown_2017] processed by the knitr package [@knitr_2017]. Another feature of MAREA is that the user can choose to share the data. Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data are saved into a central repository. These data can be accessed at any time by any person interested in acquiring them at [github.com/turfeffect/MAREAdata](github.com/turfeffect/MAREAdata).

@@

**Fig. 3. General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).

```{r map}
proj <- CRS("+proj=longlat +datum=NAD27")

@@

study_area <- tm_shape(baja2, bbox = bb(xlim = c(-115.22, -115.14), ylim = c(27.85, 27.90))) +
  tm_polygons() +
  tm_shape(reserves) +
  tm_polygons(border.col = "black", col = "red", border.lwd = 0) +
  tm_shape(controls) +
  tm_symbols(shape = 22, size = 4, border.col = "black", col = "blue", border.lwd = 1) +
  tm_text("Label") +
  tm_shape(labels2) +
  tm_text("Label") +
  tm_compass(position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom"))
```

@@

<div id="refs"></div>
